---
title: "Data Visualizations Examples"
author: "Rob Spenceley"
date: "2025-10-28"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    toc_depth: 3
---

#Animated Bar Graphs
##Pokemon Types
```{r}
library(tidyverse)
library(gganimate)
library(magick)
library(scales)
```

### --- Load + normalize ---
```{r}
df_raw <- read_csv("All_Pokemon.csv", show_col_types = FALSE)

type_candidates <- c("Type 1","Type.1","type_1","type1","Type1","PrimaryType","primary_type")
type_col <- intersect(type_candidates, names(df_raw))[1]
stopifnot(!is.na(type_col))

df <- df_raw %>%
  rename(Type1 = all_of(type_col)) %>%
  rename_with(~"Generation", .cols = matches("^Gen(eration)?$", ignore.case = TRUE)) %>%
  select(Name = matches("Name", ignore.case = TRUE), Generation, Type1) %>%
  drop_na(Generation, Type1) %>%
  mutate(Generation = as.integer(Generation),
         Type1 = as.character(Type1))
```


### --- Colors ---
```{r}
pokemon_colors <- c(
  "Normal"="#A8A77A","Fire"="#EE8130","Water"="#6390F0","Grass"="#7AC74C",
  "Electric"="#F7D02C","Ice"="#96D9D6","Fighting"="#C22E28","Poison"="#A33EA1",
  "Ground"="#E2BF65","Flying"="#A98FF3","Psychic"="#F95587","Bug"="#A6B91A",
  "Rock"="#B6A136","Ghost"="#735797","Dragon"="#6F35FC","Steel"="#B7B7CE",
  "Dark"="#705746","Fairy"="#DDA0DD"
)
type_levels <- names(pokemon_colors)
```

### --- Cumulative counts (all types every gen) ---
```{r}
cum_counts <- df %>%
  count(Generation, Type1, name = "Count") %>%
  mutate(Type1 = factor(Type1, levels = type_levels)) %>%
  complete(Type1 = factor(type_levels, levels = type_levels),
           Generation = seq(min(Generation), max(Generation)),
           fill = list(Count = 0)) %>%
  arrange(Type1, Generation) %>%
  group_by(Type1) %>%
  mutate(Cum = cumsum(Count)) %>%
  ungroup()
```

### --- Rank all types per generation ---
```{r}
ranked <- cum_counts %>%
  group_by(Generation) %>%
  arrange(Generation, desc(Cum), .by_group = TRUE) %>%
  mutate(rank = row_number(),
         type_label = as.character(Type1),
         value_label = comma(Cum)) %>%
  ungroup()

max_cum <- max(ranked$Cum, na.rm = TRUE)
left_inset  <- 0.10 * max_cum
right_inset <- 0.05 * max_cum
```

### --- Plot ---
```{r}
p <- ggplot(ranked, aes(x = rank, y = Cum, fill = Type1, group = Type1)) +
  geom_col(width = 0.80, color = NA) +
  # left-aligned type names with matching colors
  geom_text(aes(y = -left_inset,
                label = ifelse(Cum > 0, type_label, ""),
                colour = Type1),
            hjust = 1, vjust = 0.5, size = 4.6, fontface = "bold",
            lineheight = 0.9, check_overlap = TRUE) +
  # right value labels
  geom_text(aes(y = Cum + right_inset,
                label = ifelse(Cum > 0, value_label, ""),
                colour = Type1),
            hjust = 0, vjust = 0.5, size = 4, fontface = "bold",
            check_overlap = TRUE) +
  scale_fill_manual(values = pokemon_colors, drop = FALSE) +
  scale_colour_manual(values = pokemon_colors, drop = FALSE) +
  coord_flip(clip = "off") +
  scale_x_reverse() +
  scale_y_continuous(
    limits = c(-left_inset, max_cum * 1.12),
    labels = comma,
    expand = c(0, 0)
  ) +
  labs(
    title = "Cumulative Pokémon by Primary Type",
    subtitle = "All types • Up to Generation {closest_state}",
    x = NULL, y = NULL,
    caption = "Counts accumulate across generations"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "#e9eef2"),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white",  color = NA),
    plot.title = element_text(face = "bold", size = 22),
    plot.subtitle = element_text(size = 14, colour = "#555555"),
    plot.caption = element_text(size = 10, colour = "#777777"),
    plot.margin = margin(16, 36, 16, 96)
  ) +
  transition_states(Generation, transition_length = 4, state_length = 1) +
  view_follow(fixed_y = TRUE) +
  ease_aes("cubic-in-out")

# --- Animate inline (HTML) ---
print(animate(p, nframes = 140, fps = 20, renderer = magick_renderer()))
```

# MAPS
```{r}

library(tidyverse)
library(data.table)
library(osmdata)
library(sf)
library(ggrepel)
library(maps)
library(mapproj)
library(usmap)
```


#Reading the data in
```{r}
braking <- read.csv("braking_data.csv")
```

#Taking a look 
```{r}
head(braking)

view(braking)
```

# Read in the map data - Wisconsin
```{r}
wi_map_county = st_as_sf(map('county', c("wisconsin", "illinois", "minnesota"), plot = F, fill = T))
wi_only_map = st_as_sf(map('county', "wisconsin", plot = F, fill = T))

#st_as_sf is the standard way to import this kind of map. So, what you are doing here is 
#if you want to add more states, then you just change "wisconsin" to c("list of states),
#so, for the above example, you could change the county to just wisconsin if you wanted only wisconsin

view(wi_map_county)
```

the geoms here are multipolygons, so you want to wait to combine until the very end

# plot
```{r}
ggplot()+
  geom_sf(data = wi_map_county)+
  geom_point(data = braking, aes(x= lon, y = lat))+
  theme_void()

#when you are doing this kind of map, you should do your sf layer FIRST 
#For maps, USE THEME VOID!!!!!
```

# oly wisc
```{r}
ggplot()+
  geom_sf(data = wi_only_map)+
  geom_point(data = braking, aes(x= lon, y = lat))+
  theme_void()

#when you are doing this kind of map, you should do your sf layer FIRST 
#For maps, USE THEME VOID!!!!!
```

# Filter to include only Dane country 
```{r}
latitudes <- c(43.3, 42.9)
longitude <- c(-89.9, -89.0)

dan_co_hb_df = braking %>%
  filter(lat <= 43.3 & lat >= 42.9 &
           lon >= -89.9 & lon <= -89.0)

```

#reanme
```{r}
wi_map_county = st_as_sf(map('county', c("wisconsin", "illinois", "minnesota"), plot = F, fill = T))
```

# re plot

```{r}

ggplot()+
  geom_sf(data = wi_only_map)+
  geom_point(data = dan_co_hb_df, aes(x= lon, y = lat))+
  theme_void()
```

# Open street map data integration
need to get the madison region using get bounding box (gbb)
```{r}
msn = getbb("Madison, Wisconsin", featuretype = "city")

msn_streets = msn %>%
  opq() %>%
  add_osm_feature(key = "highway", 
                  value = c("motorway", "trunk", "primary", "secondary", "tertiary")) %>%
  osmdata_sf()
```

# plottig the street data 
```{r}
ggplot()+
  geom_sf(data = msn_streets$osm_lines, alpha = 0.8, size = 0.4, color = "grey")+
  geom_point(data = dan_co_hb_df, aes(x = lon, y = lat), size = 0.5)+
  lims(x = c(msn[1,1], msn[1,2]), y = c(msn[2,1], msn[2,2]))+
  theme_void()
```

# new layers
```{r}
ggplot()+
  geom_sf(data = msn_streets$osm_lines, alpha = 0.8, size = 0.4, color = "grey")+
  geom_point(data = dan_co_hb_df, aes(x = lon, y = lat), size = 0.5, color = "red")+
  lims(x = c(msn[1,1], msn[1,2]), y = c(msn[2,1], msn[2,2]))+
  facet_wrap(~participant)+
  theme_void()
```
#creating a new representation
```{r}
hb_per_part <- dan_co_hb_df %>%
  group_by(participant) %>%
  summarise(AF_count = sum(str_count(participant, "AF(0[1-9]|1[0-9]|20|21)")))

hb_per_part
```

# REAL suggestion 
```{r}
brake_sm_df = braking %>%
  group_by(participant) %>%
  summarize(total = n())

brake_sm_df
```
#Plot
this plot is less engaging but far more explainatory 
```{r}
ggplot(data = brake_sm_df, aes(x = total,xend = 0, y = reorder(participant, total),yend=reorder(participant, total)))+
  geom_point()+
  geom_segment()+
  scale_x_continuous(expand = c(0,0.), limits = c(0, 1250))+
  xlab("Total hard braking events")+
  ylab("Participant")+
 theme_bw() +
   theme(
     axis.ticks.y = element_blank(),
     axis.line.x  = element_blank(),
     panel.grid.major.y = element_blank(),
     panel.grid.minor.y = element_blank()) #this has to go after theme_bw
```


