---
title: "10.30 Class Notes"
author: "Rob Spenceley"
date: "2025-10-30"
output: html_document
---
```{r}
library(tidyverse)
library(patchwork)
library(ggplot2)
```

```{r}
mean_change_df = read_csv("simulation_results.csv")

mean_change_df = mean_change_df %>% 
  group_by(iteration, condition_mean) %>% 
  mutate(minsep_time = min(min(timestep[action == 1]),20),
         action = ifelse(timestep >= minsep_time, "Automation", "Manual"))

distr_df = read_csv("distr_data.csv")
```
# Variables

iteration - simulation iteration 100 per setting condition_mean - mean speed where automation works best condition_sd - sd of speed where automation works best timestep - time step in the simulation true_state - 0 if manual is best, 1 if automation is best speed - observed speed at a given time step (kph) belief_m - driver’s belief that manual driving is best (p 0-1) belief_a - driver’s belief that automated driving is best (p 0-1) action - whether the driver engages the automation or not

```{r}
belief_plots = ggplot()+
  geom_vline(xintercept = 10, color = "black", linetype = 2)+
  geom_path(data = mean_change_df, 
            aes(x = timestep, y = belief_m, group = iteration), 
            color = "#33a02c",
            alpha = 0.25)+
  geom_path(data = mean_change_df, 
            aes(x = timestep, y = belief_a, group = iteration), 
            color = "#1f78b4",
            alpha = 0.25)+
  facet_grid(condition_mean~.)+
  xlab("Time (Seconds)")+
  ylab("Belief in Automation (%)")+
  scale_y_continuous( 
    breaks = c(.25, .75),# <- forces those ticks to appear          # <- optional: removes padding
  )+ 
  theme_bw()

belief_plots
```

```{r}
action_plots <- ggplot(filter(mean_change_df, timestep == minsep_time),
       aes(x = minsep_time)) +
  geom_vline(xintercept = 10, color = "grey", linetype = 2) +
  geom_histogram(binwidth = 1) +
  facet_grid(condition_mean ~ .) +
  labs(x = "Time of First Automation Activation (Seconds)",
       y = "Number of Participants") +
  scale_y_continuous(
    breaks = c(0, 20, 40, 60),
    limits = c(0, 50),         # <- forces those ticks to appear
    expand = c(0, 0)           # <- optional: removes padding
  )+
  theme_bw()

# Make sure it renders:
action_plots
```

```{r}
distr_plots = ggplot(distr_df, aes(x = values, fill = condition_type))+
  stat_function(fun = dnorm, args = list(mean = 70, sd = 15), color = "#1f78b4")+
  stat_function(fun = dnorm, args = list(mean = 90, sd = 5),color = "black")+
  stat_density(alpha = 0.75, position = "dodge")+
  theme_bw()+
  scale_fill_manual(values = c("#33a02c", "#1f78b4"))+
  guides()+
  facet_grid(condition_mean~., scales = "free")+
  labs(
    x = "Belief in driving ability",
    y = "Density of believed ability",
    #title = "Belief in manual or automated driving ability at some speed (kmh)",
    fill = "Condition Type")+
  theme(legend.position = "bottom")+
  scale_y_continuous(
    breaks = c(0, .05, .12),
    expand = c(0,0)
  )
 
distr_plots
```

```{r}
distr_plots+belief_plots+action_plots +
  ggtitle("Driver Beliefs and Automation Engagement Over Time")+
  theme(plot.title = element_text(hjust = 1.3))

#this should ahve a title 
# we are trying to simplify, so a lot of what we are doing is try to simplify 
```
the green is belief in automation, blue is belive in self, so you can see that there is a swap in beleifs around time 10, which is when the distributions start to separate. So drivers that have a good understanding of what is happening is that drivers that have an accurate represendation of the world, engage the automation when they should, which is around the 10 mark. 


So this is the baseline code, and we need to adjust the codes 
you can remove the facet labels
for the assignment, we want things to match up well wuth how we did this assignment 

so, for ERPs, we can graph a "grand average" ERP 

/n allows you to use a new line