---
title: "10.23 class notes"
author: "Rob Spenceley"
date: "2025-10-23"
output: html_document
---


#Loading packages
```{r}
#you should do all of this in this order so you make sure you are calling the right functions 

library(tidyverse)
library(data.table)
library(osmdata)
library(sf)
library(ggrepel)
library(maps)
library(mapproj)
library(usmap)

set.seed(48)
```

#Reading the data in
```{r}
braking <- read.csv("braking_data.csv")
```

#Taking a look 
```{r}
head(braking)

view(braking)
```
acc_br= braking 
acc_sw = sway 
so this is showing data from a study where there was hard braking incidents. So if braking acceleration is over a certain threshold then this would be considered a hard braking event
so every point in this data is a hard braking event, and we have the coordinates of that data, so, we want to see where it is occuring 

# Read in the map data - Wisconsin
```{r}
wi_map_county = st_as_sf(map('county', c("wisconsin", "illinois", "minnesota"), plot = F, fill = T))
wi_only_map = st_as_sf(map('county', "wisconsin", plot = F, fill = T))

#st_as_sf is the standard way to import this kind of map. So, what you are doing here is 
#if you want to add more states, then you just change "wisconsin" to c("list of states),

view(wi_map_county)
```

the geoms here are multipolygons, so you want to wait to combine until the very end

# plot
```{r}
ggplot()+
  geom_sf(data = wi_map_county)+
  geom_point(data = braking, aes(x= lon, y = lat))+
  theme_void()

#when you are doing this kind of map, you should do your sf layer FIRST 
#For maps, USE THEME VOID!!!!!
```

# oly wisc
```{r}
ggplot()+
  geom_sf(data = wi_only_map)+
  geom_point(data = braking, aes(x= lon, y = lat))+
  theme_void()

#when you are doing this kind of map, you should do your sf layer FIRST 
#For maps, USE THEME VOID!!!!!
```

# Filter to include only Dane country 
```{r}
latitudes <- c(43.3, 42.9)
longitude <- c(-89.9, -89.0)

dan_co_hb_df = braking %>%
  filter(lat <= 43.3 & lat >= 42.9 &
           lon >= -89.9 & lon <= -89.0)

```

#reanme
```{r}
wi_map_county = st_as_sf(map('county', c("wisconsin", "illinois", "minnesota"), plot = F, fill = T))
```

# re plot

```{r}

ggplot()+
  geom_sf(data = wi_only_map)+
  geom_point(data = dan_co_hb_df, aes(x= lon, y = lat))+
  theme_void()
```

# Open street map data integration
need to get the madison region using get bounding box (gbb)
```{r}
msn = getbb("Madison, Wisconsin", featuretype = "city")

msn_streets = msn %>%
  opq() %>%
  add_osm_feature(key = "highway", 
                  value = c("motorway", "trunk", "primary", "secondary", "tertiary")) %>%
  osmdata_sf()
```

# plottig the street data 
```{r}
ggplot()+
  geom_sf(data = msn_streets$osm_lines, alpha = 0.8, size = 0.4, color = "grey")+
  geom_point(data = dan_co_hb_df, aes(x = lon, y = lat), size = 0.5)+
  lims(x = c(msn[1,1], msn[1,2]), y = c(msn[2,1], msn[2,2]))+
  theme_void()
```

# new layers
```{r}
ggplot()+
  geom_sf(data = msn_streets$osm_lines, alpha = 0.8, size = 0.4, color = "grey")+
  geom_point(data = dan_co_hb_df, aes(x = lon, y = lat), size = 0.5)+
  lims(x = c(msn[1,1], msn[1,2]), y = c(msn[2,1], msn[2,2]))+
  facet_wrap(~participant)+
  theme_void()
```
#creating a new representation
```{r}
hb_per_part <- dan_co_hb_df %>%
  group_by(participant) %>%
  summarise(AF_count = sum(str_count(participant, "AF(0[1-9]|1[0-9]|20|21)")))

hb_per_part
```
#plotting totals
```{r}
ggplot(hb_per_part, aes(x = participant, y = AF_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(x = "Participant", y = "AF Count", title = "AF01–AF21 Count per Participant") +
  coord_flip()+
  theme_bw()
```

# REAL suggestion 
```{r}
brake_sm_df = braking %>%
  group_by(participant) %>%
  summarize(total = n())

brake_sm_df
```
#Plot
this plot is less engaging but far more explainatory 
```{r}
ggplot(data = brake_sm_df, aes(x = total,xend = 0, y = reorder(participant, total),yend=reorder(participant, total)))+
  geom_point()+
  geom_segment()+
  scale_x_continuous(expand = c(0,0.), limits = c(0, 1250))+
  xlab("Total hard braking events")+
  ylab("Participant")+
 theme_bw() +
   theme(
     axis.ticks.y = element_blank(),
     axis.line.x  = element_blank(),
     panel.grid.major.y = element_blank(),
     panel.grid.minor.y = element_blank()) #this has to go after theme_bw
```

